A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-18 tex/body.tex
some edits

\newcommand\SSS{\mathit{S\vthinspace}}
\newcommand\SSSS[1]{\mathit{S}^{\,#1}}

@@

The decision procedure for (co)datatypes and its extension to quantified
formulas were described at a high level of abstraction. This section describes
the main aspects of the implementation within the SMT solver CVC4:\ the
integration into CDCL($T$), the extension to quantified formulas, and
some optimizations.

@@

%\paragraph{Integration into {\relax\XXXL($T\negvthinspace$)}.}
The calculus is implemented as a \emph{theory solver} of CVC4, that is, a
specialized solver for determining the satisfiability of conjunctions of
constraints for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and an input set of
clauses $F$ in CNF, the \XXXL($T$) procedure
incrementally builds partial assignments from the atomic formulas of $F$ to
truth values such that no clause in $F$ is falsified. We may interpret such a
partial assignment as a set of literals where $a \in M$ for all atoms $a$ that
it assigned the value $\top$, and $\neg\; a \in M$ for all atoms that it assigns
the value $\bot$. Then, at a high level, by the Nelson--Oppen combination
framework (its variant in CVC4 is described in~\cite{jovanovic2011sharing}), 
each $T_i$-solver for $i = 1, \ldots, l$ takes as input a combination
of (1) the purified form of $T_i$-constraints occurring in $M$, where terms
containing symbols not belonging to $T_i$ are replaced by fresh variables of the
same type; (2) additional equalities and disequalities between variables of
types not belonging to $T_i$. Let us call this set $M_i$. Each $T_i$ solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\; C$ is added to $F$, adds a clause to $F$, or does nothing.
When $M$ is a complete assignment for $F$, a theory solver can only choose to do
nothing when $M_i$ is indeed $T_i$-satisfiable.

Assume $\Ec$ is initially the set $M_i$ described above. For each equality $t \teq s$ added
to $\Ec$, we associate a set of equalities from $M_i$ that together entail $t
\teq s$, which we call its \emph{explanation}.
Similarly, each $\Val \ec{x}$ is also associated an explanation---that is, a set
of equalities from $M_i$ that entail that the values of $\ec{x}$ in models of
$\Ec$ are of the form $\Val \ec{x}$. For example, if $x \teq \const{C}( x ) \in
M_i$, then $x \teq \const{C}( x )$ is an %(possible)
explanation for $\Val \ec{x} = \MU {\Varec{x}}.\; \const{C}( {\Varec{x}} )$.
Thus, the rules of the calculus are implemented as follows. For all rules with
conclusion $\bot$, we report the union of the explanations for all premises is
$\thD$-unsatisfiable. For \rn{Split}, we add a clause %of the form
$t \teq \const{C}_1\bigl( \const s^1_1( t ), \ldots, \const s^{n_{1\!}}_1( t ) \bigr)
\mathrel\lor \cdots \mathrel\lor t \teq \const{C}_m\bigl( \const s^1_m( t ),
\ldots, \const s^{n_m}_m( t ) \bigr)$ to $F$.
Decisions on which branch to take are thus performed externally by the SAT
solver. % and not by the theory solver.
All other rules add equalities to the
internal state of the theory solver. The rules in phase~1 are performed
eagerly---that is, for partial satisfying assignments $M$---while the rules in
phases~2 and~3 are performed only for complete satisfying
assignments $M$.

Prior to constructing a model for $F$, the theory solver does not explicitly construct $\mu$-terms nor the map $\Val$.
Instead, $\Val$ is computed implicitly by traversing the equivalence classes of $\Ec$ during phase~2.
To recognize when the precondition of \rn{Acyclic} holds,
the procedure considers each datatype equivalence class $\ec{t}$ containing a constructor $\const{C}( t_1, \ldots, t_n )$.
It visits $\ec{t_1}, \ldots, \ec{t_n}$ and all arguments of constructors in these equivalence classes in the same way.
If while doing so it returns to $\ec{t}$, then it deduces that \rn{Acyclic} applies to $t$.
To recognize when the precondition of \rn{Unique} holds,
the procedure considers the set $S$ of all codatatype equivalence classes.
It simultaneously visits the equivalence classes of arguments of constructor terms in each equivalence class in $S$,
while partitioning $S$ into $S_1, \ldots, S_n$ based on the top-most symbol of constructor terms in these equivalence classes
and the equivalence of their arguments of type $\Nondata$.
It then partitions each set in this partition in the same manner recursively.
If after doing so, the resulting partition contains a set $S_i$ containing distinct terms $u$ and $v$, then it deduces that \rn{Unique} applies to $u$ and $v$.

>>>>

\newcommand\SSS{\mathit{S}}
\newcommand\SSSS[1]{\mathit{S}^{\,#1}}

@@

The decision procedure for (co)datatypes and its extension to quantified
formulas were presented at a high level of abstraction. This section describes
the main aspects of the implementation within the SMT solver CVC4:\ the
integration into CDCL($T$), the extension to quantified formulas, and
some optimizations.

@@

%\paragraph{Integration into {\relax\XXXL($T\negvthinspace$)}.}
The calculus is implemented as a \relax{theory solver} of CVC4, that is, a
specialized solver for determining the satisfiability of conjunctions of
constraints for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and a set of
input clauses $F$ in CNF, the \XXXL($T$) procedure
incrementally builds partial assignments from the atomic formulas of $F$ to
truth values such that no clause in $F$ is falsified. We can interpret such a
partial assignment as a set $M$ of literals where $a \in M$ for all atoms $a$ that
it assigned the value $\top$ and $\neg\; a \in M$ for all atoms that it assigns
the value $\bot$. Then, %at a high level,
by a variant
\cite{jovanovic2011sharing} of the Nelson--Oppen method,
each $T_i$-solver for $i = 1, \ldots, l$ takes as input a combination
of (1) the purified form of $T_i$-constraints occurring in $M$, where terms
containing symbols not belonging to $T_i$ are replaced by fresh variables \begin{rep}of the
same type\end{rep}; (2) additional equalities and disequalities between variables of
types not belonging to $T_i$. Let us call this set $M_i$. Each $T_i$ solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\; C$ is added to $F\!$, adds a clause to $F\!$, or does nothing.
When $M$ is a complete assignment for $F\!$, a theory solver can only choose to do
nothing when $M_i$ is indeed $T_i$-satisfiable.

Assume $\Ec$ is initially the set $M_i$ described above. With each equality $t \teq s$ added
to $\Ec$, we associate a set of equalities from $M_i$ that together entail $t
\teq s$, which we call its \emph{explanation}.
Similarly, each $\Val \ec{x}$ is assigned an explanation---that is, a set
of equalities from $M_i$ that entail that the values of $\ec{x}$ in models of
$\Ec$ are of the form $\Val \ec{x}$. For example, if $x \teq \const{C}( x ) \in
M_i$, then $x \teq \const{C}( x )$ is an %(possible)
explanation for $\Val \ec{x} = \MU {\Varec{x}}.\; \const{C}( {\Varec{x}} )$.
Thus, the rules of the calculus are implemented as follows. For all rules with
conclusion $\bot$, we report the union of the explanations for all premises is
$\thD$-unsatisfiable. For \rn{Split}, we add a clause %of the form
$t \teq \const{C}_1\bigl( \const s^1_1( t ), \ldots, \const s^{n_{1\!}}_1( t ) \bigr)
\mathrel\lor \cdots \mathrel\lor t \teq \const{C}_m\bigl( \const s^1_m( t ),
\ldots, \const s^{n_m}_m( t ) \bigr)$ to $F$.
Decisions on which branch to take are thus performed externally by the SAT
solver. % and not by the theory solver.
All other rules add equalities to the
internal state of the theory solver. The rules in phase~1 are performed
eagerly---that is, for partial satisfying assignments $M$---while the rules in
phases~2 and~3 are performed only for complete satisfying
assignments $M$.

Prior to constructing a model for $F\!$, the theory solver does not explicitly construct $\mu$-terms nor the mapping $\Val$.
Instead, $\Val$ is computed implicitly by traversing the equivalence classes of $\Ec$ during phase~2.
To recognize when the precondition of \rn{Acyclic} holds,
the procedure considers each equivalence class $\ec{t}$ containing a datatype constructor $\const{C}( t_1, \ldots, t_n )$.
It visits $\ec{t_1}, \ldots, \ec{t_n}$ and all arguments of constructors in these equivalence classes recursively.
If while doing so it returns to $\ec{t}$, then it deduces that \rn{Acyclic} applies to $t$.
To recognize when the precondition of \rn{Unique} holds,
the procedure considers the set $\SSS$ of all codatatype equivalence classes.
It simultaneously visits the equivalence classes of arguments of constructor terms in each equivalence class in $\SSS$,
while partitioning $\SSS$ into $\SSS_1, \ldots, \SSS_n$ based on the top-most symbol of constructor terms in these equivalence classes
and the equivalence of their arguments of type $\Nondata$.
It then partitions each set in this partition in the same manner recursively.
If %after doing so,
the resulting partition contains a set $\SSS_i$ containing distinct terms $u$
and $v$, it deduces that \rn{Unique} applies to $u$ and $v$.
